---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BioAge

This package measures biological aging using data from the National Health and Nutrition Examination Survey (NHANES). The package uses published biomarker algorithms to calculate three biological aging measures: Klemera-Doubal Method (KDM) biological age, phenotypic age, and homeostatic dysregulation.

## Installation (via devtools):

You can install the released version of BioAge from (https://github.com/dayoonkwon/BioAge) with:

``` r
install.packages("devtools")
devtools::install_github("dayoonkwon/BioAge")
```
## Example

This serves as an example of training biologial aging measures using the NHANES 3 (1991) and NHANES 4 (1999 - 2018) dataset.
It also provides documentation for fit parameters contained in the ```BioAge``` package.
The cleaned NHANES dataset is loaded as the dataset ```NHANES3``` and ```NHANES4```.
The original KDM bioage and phenoage values are saved as ```kdm0``` and ```phenoage0``` as part of NHANES dataset. 

```{r setup}
library(BioAge) #topic of example
library(dplyr)
```

## Step 1: train in NHANES 3 and project in NHANES 4

I train in the NHANES 3 and project biological aging measures into the NHANES 4 by using the ```hd_nhanes```, ```kdm_nhanes```, and ```phenoage_nhanes``` function of the ```BioAge``` package.

```{r}
#particular biomarkers
biomarkers = c("albumin","alp","lymph","mcv","lncreat","lncrp","hba1c","wbc","rdw")

#projecting HD using NHANES (seperate training for gender)
hd = hd_nhanes(biomarkers)

#projecting KDM bioage using NHANES (seperate training for gender)
kdm = kdm_nhanes(biomarkers)

#projecting phenoage uinsg NHANES
phenoage = phenoage_nhanes(biomarkers)
```

