---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BioAge

This package measures biological aging using data from the National Health and Nutrition Examination Survey (NHANES). The package uses published biomarker algorithms to calculate three biological aging measures: Klemera-Doubal Method (KDM) biological age, phenotypic age, and homeostatic dysregulation.

## Installation (via devtools):

You can install the released version of BioAge from (https://github.com/dayoonkwon/BioAge) with:

``` r
# install.packages("devtools")
devtools::install_github("dayoonkwon/BioAge")
```
## Example

This serves as an example of training biologial aging measures using the NHANES 3 (1991) and NHANES 4 (1999 - 2018) dataset.
It also provides documentation for fit parameters contained in the BioAge package.
The cleaned NHANES dataset is loaded as the dataset ```NHANES3``` and ```NHANES4```.
The original KDM bioage and phenoage values are saved as ```kdm0``` and ```phenoage0``` as part of NHANES dataset. 

```{r setup}
library(dplyr)
library(BioAge) #topic of vignette
```

## Step 1: train in NHANES 3 and project in NHANES 4

I train in the NHANES 3 and project biological aging measures into the NHANES 4 by using the ```hd_nhanes```, ```kdm_nhanes```, and ```phenoage_nhanes``` function of the BioAge package.

```{r}
#particular biomarkers
biomarkers = c("albumin","alp","lymph","mcv","lncreat","lncrp","hba1c","wbc","rdw")

#projecting HD using NHANES (seperate training for gender)
hd = hd_nhanes(biomarkers)

#projecting KDM bioage using NHANES (seperate training for gender)
kdm = kdm_nhanes(biomarkers)

#projecting phenoage uinsg NHANES
phenoage = phenoage_nhanes(biomarkers)
```

## Step 2: compare NHANES 4 to the original KDM bioage and phenoage

The projected data and estimated model above are saved as part of the list structure.
These can be drawn by typing ```data``` and ```fit```, respectively.

```{r}
#pull the full dataset
data = merge(hd$data, kdm$data) %>% merge(., phenoage$data)
```

### Figure1: Chronological age vs biological aging measures

```{r, fig.width = 8, fig.height = 7}
#select biological age variables
agevar = c("kdm0","phenoage0","kdm","phenoage","hd","hd_log")

#prepare labels
label = c("KDM\nBiological Age",
          "Levine\nPhenotypic Age",
          "Modified-KDM\nBiological Age",
          "Modified-Levine\nPhenotypic Age",
          "Mahalanobis\nDistance",
          "Log\nMahalanobis\nDistance")

#plot age vs bioage
plot_ba(data, agevar, label)

```

### Figure2: Corplot for biological aging measures

```{r, fig.width = 8, fig.height = 8}
#select biological age advancement (BAA) variables
agevar = c("kdm_advance0","phenoage_advance0","kdm_advance","phenoage_advance","hd","hd_log")

#prepare lables
#values should be formatted for displaying along diagonal of the plot
#names should be used to match variables and order is preserved
label = c(
  "kdm_advance0"="KDM\nBiological\nAge",
  "phenoage_advance0"="Levine\nPhenotypic\nAge",
  "kdm_advance"="Modified\nKDM\nBiological\nAge",
  "phenoage_advance"="Modified\nLevine\nPhenotypic\nAge",
  "hd" = "Mahalanobis\nDistance",
  "hd_log" = "Log\nMahalanobis\nDistance")

#use variable name to define the axis type ("int" or "float")
axis_type = c(
  "kdm_advance0"="float",
  "phenoage_advance0"="float",
  "kdm_advance"="float",
  "phenoage_advance"="flot",
  "hd"="flot",
  "hd_log"="float")

#plot BAA corplot
plot_baa(data,agevar,label,axis_type)
```

### Table 1: Mortality models with all biological aging measures

```{r}
table_surv(data, agevar, label)
```

