---
title: "Using NHANES to Train and Project into HRS"
author: "Dayoon Kwon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Overview

This vignette serves as an example of training biologial aging measures using the NHANES 3 (1991) and NHANES 4 (1999 - 2018) dataset.
It also provides documentation for fit parameters contained in the BioAge package.
The cleaned NHANES dataset is loaded as the dataset ```NHANES3``` and ```NHANES4```.
The original KDM bioage and phenoage values are saved as ```kdm0``` and ```phenoage0``` as part of NHANES dataset. 

## Installation (via devtools):

You can install the released version of BioAge from (https://github.com/dayoonkwon/BioAge) with:

```{r setup}
library(dplyr)
library(BioAge) #topic of vignette
```

## Step 1: train in NHANES 3 and project in NHANES 4

I train in the NHANES 3 and project biological aging measures into the NHANES 4 by using the ```hd_nhanes```, ```kdm_nhanes```, and ```phenoage_nhanes``` function of the BioAge package.

```{r}
#particular biomarkers
biomarkers = c("albumin","alp","lymph","mcv","lncreat","lncrp","hba1c","wbc","rdw")

#projecting HD using NHANES (seperate training for gender)
hd = hd_nhanes(biomarkers)

#projecting KDM bioage using NHANES (seperate training for gender)
kdm = kdm_nhanes(biomarkers)

#projecting phenoage uinsg NHANES
phenoage = phenoage_nhanes(biomarkers)
```

## Step 2: compare NHANES 4 to the original KDM bioage and phenoage

The projected data and estimated model above are saved as part of the list structure.
These can be drawn by typing ```data``` and ```fit```, respectively.

```{r}
#pull the full dataset
data = merge(hd$data, kdm$data) %>% merge(., phenoage$data)
```

### Figure1: Chronological age vs biological aging measures

```{r, fig.width = 8, fig.height = 7}
#select biological age variables
agevar = c("kdm0","phenoage0","kdm","phenoage","hd","hd_log")

#prepare labels
label = c("KDM\nBiological\nAge",
          "Levine\nPhenotypic\nAge",
          "Modified-KDM\nBiological\nAge",
          "Modified-Levine\nPhenotypic\nAge",
          "Mahalanobis\nDistance",
          "Log\nMahalanobis\nDistance")

#plot age vs bioage
plot_ba(data, agevar, label)

```

### Figure2: Corplot for biological aging measures

```{r, fig.width = 8, fig.height = 8}
#select biological age advancement (BAA) variables
agevar = c("kdm_advance0","phenoage_advance0","kdm_advance","phenoage_advance","hd","hd_log")

#prepare lables
#values should be formatted for displaying along diagonal of the plot
#names should be used to match variables and order is preserved
label = c(
  "kdm_advance0"="KDM\nBiological\nAge",
  "phenoage_advance0"="Levine\nPhenotypic\nAge",
  "kdm_advance"="Modified\nKDM\nBiological\nAge",
  "phenoage_advance"="Modified\nLevine\nPhenotypic\nAge",
  "hd" = "Mahalanobis\nDistance",
  "hd_log" = "Log\nMahalanobis\nDistance")

#use variable name to define the axis type ("int" or "float")
axis_type = c(
  "kdm_advance0"="float",
  "phenoage_advance0"="float",
  "kdm_advance"="float",
  "phenoage_advance"="flot",
  "hd"="flot",
  "hd_log"="float")

#plot BAA corplot
plot_baa(data,agevar,label,axis_type)
```

### Table 1: Mortality models with all biological aging measures

```{r}
#creat table for mortality models
table_surv(data, agevar, label)
```

### Table 2: Linear regression models with current health status outcomes

```{r}
table2 = table_health(data,agevar,outcome = c("health","adl","lnwalk","grip_scaled"), label)
table2$table
table2$n
```

### Table 3: Linear regresion models with socioeconomic variables

```{r}
table3 = table_ses(data,agevar,exposure = c("edu","annual_income","poverty_ratio"), label)
table3$table
table3$n
```

## Step 3: Score new data
The projection dataset is the Health and Retirement Study (HRS), which has identical biomarkers, and was previously cleanded.
I train in the NHANES 3 and project biological aging measures into the HRS by using the ```hd_calc```, ```kdm_calc```, and ```phenoage_calc``` function of the BioAge package.

```{r, echo=FALSE}
#The HRS dataset is loaded from my local drive that has previously been downloaded and cleaned
#The original data is not available.
newdata = readstata13::read.dta13("/Users/dayoonkwon/Dropbox/belskylab/project/hrs/HRS/HRS_DWB_forBA.dta") %>%
  select(hhidpn, sex, age, raracem, adls, grip, srh, lnwalk,
         albumin, alp, lymphpct, mcv, creat, lncreat, crp, lncrp, hba1c, wbc, rdw) %>%
  rename(sampleID = hhidpn,
         gender = sex,
         race = raracem,
         adl = adls,
         grip_scaled = grip,
         health = srh,
         lymph = lymphpct) %>%
  mutate(gender = ifelse(gender == "Women", 2,
                         ifelse(gender == "Men", 1, NA)),
         albumin_gL = albumin * 10,
         creat_umol = creat * 88.4017,
         lncreat_umol = log(creat_umol),
         crp = crp / 10,
         lncrp = log(crp)) %>%
  group_by(gender) %>%
  mutate_at(vars(albumin:creat,crp,hba1c:rdw),
            list(~ifelse((. > (mean(., na.rm = TRUE) + 5 * sd(., na.rm = TRUE)))|
                           (. < (mean(., na.rm = TRUE) - 5 * sd(., na.rm = TRUE))), NA, .))) %>%
  ungroup() %>%
  mutate(lncreat = ifelse(is.na(creat), NA, lncreat),
         lncrp = ifelse(is.na(crp), NA, lncrp),
         albumin_gL = ifelse(is.na(albumin), NA, albumin_gL),
         creat_umol = ifelse(is.na(creat), NA, creat_umol),
         lncreat_umol = ifelse(is.na(creat), NA, lncreat_umol))
```

For HD, the constructed varialbe is based on a malhanobis distance statistic, which is theoretically the distance between observations and a hypothetically healthy, young cohort.
In this example, I train separately for men and women who are between the ages of 20 and 30 and not pregnant, and have observe biomarker data within clinically accpetable distributions.

```{r}
#projecting HD  into the HRS using NHANES 3 (seperate training for gender)
hd_fem = hd_calc(data = newdata %>%
                   filter(gender == 2),
                 reference = NHANES3 %>%
                   filter(gender == 2 & age >= 20 & age <= 30 & pregnant == 0) %>%
                   mutate(albumin = ifelse(albumin >= 3.5 & albumin <= 5, albumin, NA),
                          alp = ifelse(alp >= 37 & alp <= 98, alp, NA),
                          lymph = ifelse(lymph >= 20 & lymph <= 40, lymph, NA),
                          mcv = ifelse(mcv >= 78 & mcv <= 101, mcv, NA),
                          creat = ifelse(creat >= 0.6 & creat <= 1.1, creat, NA),
                          lncreat = ifelse(is.na(creat), NA, lncreat),
                          crp = ifelse(crp < 2, crp, NA),
                          lncrp = ifelse(is.na(crp), NA, lncrp),
                          hba1c = ifelse(hba1c >= 4 & hba1c <= 5.6, hba1c, NA),
                          wbc = ifelse(wbc >= 4.5 & wbc <= 11, wbc, NA),
                          rdw = ifelse(rdw >= 11.5 & rdw <= 14.5, rdw, NA)),
                 biomarkers)

hd_male = hd_calc(data = newdata %>%
                   filter(gender == 1),
                 reference = NHANES3 %>%
                   filter(gender == 1 & age >= 20 & age <= 30 & pregnant == 0) %>%
                   mutate(albumin = ifelse(albumin >= 3.5 & albumin <= 5, albumin, NA),
                          alp = ifelse(alp >= 45 & alp <= 115, alp, NA),
                          lymph = ifelse(lymph >= 20 & lymph <= 40, lymph, NA),
                          mcv = ifelse(mcv >= 82 & mcv <= 102, mcv, NA),
                          creat = ifelse(creat >= 0.8 & creat <= 1.3, creat, NA),
                          lncreat = ifelse(is.na(creat), NA, lncreat),
                          crp = ifelse(crp < 2, crp, NA),
                          lncrp = ifelse(is.na(crp), NA, lncrp),
                          hba1c = ifelse(hba1c >= 4 & hba1c <= 5.6, hba1c, NA),
                          wbc = ifelse(wbc >= 4.5 & wbc <= 11, wbc, NA),
                          rdw = ifelse(rdw >= 11.5 & rdw <= 14.5, rdw, NA)),
                 biomarkers)

#pull the HD dataset
hd_data = rbind(hd_fem$data, hd_male$data)
```

Having estimated biological aging models using NHANES 3 in Step 1, I can project KDM bioage and phenoage into the HRS data by running ```kdm_calc``` and ```phenoage_calc``` and supplying a ```fit``` argument.

```{r}
#projecting KDM bioage into the HRS using NHANES 3 (seperate training for gender)
kdm_fem = kdm_calc(data = newdata %>%
                     filter (gender ==2),
                   biomarkers,
                   fit = kdm$fit$female,
                   s_ba2 = kdm$fit$female$s_b2)

kdm_male = kdm_calc(data = newdata %>%
                     filter (gender ==1),
                   biomarkers,
                   fit = kdm$fit$male,
                   s_ba2 = kdm$fit$male$s_b2)

#pull the KDM dataset
kdm_data = rbind(kdm_fem$data, kdm_male$data)

#projecting phenoage into the HRS using NHANES 3
phenoage_hrs = phenoage_calc(data = newdata %>%
                           mutate(albumin = albumin_gL,
                                  lncreat = lncreat_umol),
                         biomarkers,
                         fit = phenoage$fit)
```
